---
title: "Measles Vaccination Rates in US Schools"
subtitle: "Tidy Tuesday 2020, Week 9"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
bibliography: ../../refs/references.bib
---

## Background

This exploration uses the Tidy Tuesday measles vaccination dataset [@tidytuesday], originally compiled by The Wall Street Journal [@wsj_measles]. The dataset covers school-level MMR (Measles, Mumps, and Rubella) and overall vaccination rates across 32 US states, along with exemption rates (religious, medical, and personal).

## Load packages and data

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(ggridges)

measles <- readr::read_csv(
 'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-02-25/measles.csv'
)
```

## First look

```{r}
#| label: glimpse

glimpse(measles)
```

```{r}
#| label: summary

summary(measles)
```

## Data cleaning

Many MMR values are coded as `-1`, which represents missing data. Let's recode these and examine what we have.

```{r}
#| label: clean

measles_clean <- measles |>
  mutate(
    mmr = if_else(mmr < 0, NA_real_, mmr),
    overall = if_else(overall < 0, NA_real_, overall)
  )

measles_clean |>
  summarise(
    n_schools = n(),
    n_states = n_distinct(state),
    mmr_available = sum(!is.na(mmr)),
    overall_available = sum(!is.na(overall)),
    pct_mmr_available = round(100 * mmr_available / n_schools, 1)
  )
```

## MMR vaccination rates by state

```{r}
#| label: fig-mmr-by-state
#| fig-cap: "Distribution of school-level MMR vaccination rates by state"
#| fig-height: 8
#| fig-width: 10

measles_clean |>
  filter(!is.na(mmr)) |>
  mutate(state = fct_reorder(state, mmr, .fun = median, na.rm = TRUE)) |>
  ggplot(aes(x = mmr, y = state)) +
  geom_boxplot(outlier.size = 0.5, fill = "steelblue", alpha = 0.5) +
  geom_vline(xintercept = 95, linetype = "dashed", colour = "red") +
  annotate("text", x = 94, y = 1, label = "95% herd immunity threshold",
           hjust = 1, colour = "red", size = 3) +
  labs(
    x = "MMR Vaccination Rate (%)",
    y = NULL,
    title = "School-level MMR vaccination rates vary widely across states",
    subtitle = "Red dashed line shows the ~95% threshold for herd immunity"
  ) +
  theme_minimal()
```

### Ridgeplot

The boxplots above highlight many outliers. A ridgeplot shows the full distribution shape for each state, making it easier to see whether low-coverage schools are isolated exceptions or part of a broader pattern.

```{r}
#| label: fig-mmr-ridgeplot
#| fig-cap: "Density ridgeplot of school-level MMR vaccination rates by state"
#| fig-height: 10
#| fig-width: 10

measles_clean |>
  filter(!is.na(mmr)) |>
  mutate(state = fct_reorder(state, mmr, .fun = median, na.rm = TRUE)) |>
  ggplot(aes(x = mmr, y = state, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis_c(option = "C", name = "MMR %") +
  geom_vline(xintercept = 95, linetype = "dashed", colour = "red") +
  labs(
    x = "MMR Vaccination Rate (%)",
    y = NULL,
    title = "Distribution of school-level MMR vaccination rates by state",
    subtitle = "Red dashed line shows the ~95% herd immunity threshold"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## School type comparison

```{r}
#| label: fig-school-type
#| fig-cap: "MMR vaccination rates by school type"

measles_clean |>
  filter(!is.na(mmr), !is.na(type), type != "") |>
  ggplot(aes(x = type, y = mmr, fill = type)) +
  geom_boxplot(alpha = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = 95, linetype = "dashed", colour = "red") +
  labs(
    x = "School Type",
    y = "MMR Vaccination Rate (%)",
    title = "How do vaccination rates differ by school type?"
  ) +
  theme_minimal()
```

## Exemption rates

```{r}
#| label: fig-exemptions
#| fig-cap: "Relationship between personal exemptions and MMR vaccination rate"

measles_clean |>
  filter(!is.na(mmr), !is.na(xper), xper >= 0) |>
  ggplot(aes(x = xper, y = mmr)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "loess", colour = "blue", se = TRUE) +
  geom_smooth(method = "lm", colour = "red", se = TRUE) +
  labs(
    x = "Personal Exemption Rate (%)",
    y = "MMR Vaccination Rate (%)",
    title = "Higher personal exemption rates are associated with lower MMR coverage"
  ) +
  theme_minimal()
```

## Low-coverage schools

Following tidy data principles [@wickham2014tidy], let's identify schools with notably low coverage.

```{r}
#| label: low-coverage

measles_clean |>
  filter(!is.na(mmr), mmr < 80) |>
  count(state, sort = TRUE) |>
  head(10) |>
  knitr::kable(
    col.names = c("State", "Schools with <80% MMR"),
    caption = "States with the most schools below 80% MMR vaccination rate"
  )
```

## References
